\hspace{1cm} Do syntactic and semantic similarity lead to interference effects? \newline Evidence from self-paced reading and event-related potentials using German

### 2024-01-19 paper/main.tex
Updated via the Open Science Framework

Reading studies have been the dominant method used to investigate retrieval interference \parencite[see the review in][]{jaeger_etal_2017}.  We aimed to investigate the role of syntactic and semantic cues during sentence comprehension with higher-powered experiments using  self-paced reading as well as electroencephalography (EEG), which to our knowledge has never been used to investigate interference during subject-verb dependency resolution.

@@

The SPR experiments were conducted web-based via Prolific and the EEG experiment was of course run in the lab.  Because the high signal-to-noise ratio of EEG data requires a large number of trials per condition and participant, we constructed 120 items. In order to keep the SPR and EEG studies comparable, we also used 120 items in the SPR experiments. In both methods, as it would have been very taxing for the participants to read all 120 items in one experimental session, we split the experiments into two sessions.

The SPR study, hereafter Experiment 1a, was initially designed so that participants read items in the first session (1-60) and then, after a gap of 1-20 days, they read the the second set of items (61-120). The variation between participants in the gap is due to the fact that we could not control when the participants would do the second session; however, the mean and median gaps were  about 11 days, with the first and third quartiles being 9 and 12 days.

@@

The Bayes factor is analogous to the frequentist ANOVA or likelihood ratio test: it compares the marginal likelihood of a full model to that of a baseline null model; this is often written as $BF_{10}$, with the subscripts indicating that marginal likelihood of the full model is in the numerator and that of the null model is in the denominator. One major advantage of the Bayes factor over the frequentist analogue is that it takes the uncertainty of the parameters into account. This leads to more conservative inferences compared to frequentist ANOVA, which only takes the maximum likelihood estimate of the parameter into account \parencite[see][for detailed discussion]{schad_etal_2022_BF}. Another important advantage of the Bayes factor -- one that is very relevant for the present work -- is that it is in principle possible to find evidence both for or against an effect. This stands in contrast to the frequentist ANOVA which, in its standard usage, is designed to only furnish evidence against the null.

@@

\subsubsection{Comparison with previous findings}
The results of the present study stand in contrast to previous findings \citep{vandyke07, mertzen}. The present study is the only one to find semantic and no syntactic interference effect. The difference in observed effects could be explained by differences in statistical power or by methodological differences. The present study has higher statistical power than the previous studies. \citet{vandyke07} had 35 - 40 participants and 36 - 48 items in each of her three experiments, respectively. \textcite{mertzen} tested 61 English speakers and 121 German speakers with 40 items each. In contrast, we tested 774 participants with at least 60 items each (a subset of 160 participants read all 120 items). Therefore, the present study is the study with the highest power to date using this 2 $\times$ 2 interference design. Consequently, the estimates from the present study are the most precise ones so far for this design. The smaller effect size in the present study is a consequence of higher power, and the magnitude of the effects observed is comparable to that of other large-scale reading studies \citep{nicenboim} and meta-analyses on interference \citep{jaeger_etal_2017}.

An obvious difference from previous studies is the experimental methodology. The majority of the previous experiments employed eye-tracking while reading, while we used a self-paced reading task. Self-paced reading (SPR) could trigger a greater tendency to cause interference effects compared to natural reading (as in eye-tracking studies), due to the restricted reading format in SPR: Comprehenders have no opportunity to revisit previous material during conventional self-paced reading \citep[but see][]{BSPR}. SPR is likely to be more demanding on the comprehender's memory, which might make it easier to detect interference effects. This argumentation is supported by our estimation of the results of \citeauthor{vandyke07}'s (\citeyear{vandyke07}) Experiment 1, the only other SPR experiment using this design: 
In the critical region, the estimates of both main effects (syntactic and semantic interference) are very similar to the results of the present study; the interaction effect observed by \citeauthor{vandyke07} in the spill-over region is also similar to our estimate. However, it remains unclear why semantic interference was especially prominent in our study compared to the previous ones.

The contrast between our results and those from  \citeauthor{mertzen}'s (\citeyear{mertzen}) German experiment is surprising given that both used German items and partially even identical items. So, why did we find a semantic interference effect in all regions, whereas \citet{mertzen} only found one in the pre-critical region? Apart from the methodological difference (eye-tracking vs.\ self-paced reading), a plausible  explanation for the divergence is lower power in \citeauthor{mertzen}'s (\citeyear{mertzen}) study. While they already had considerably more participants and items than many psycholinguistic studies, in the present study the number of participants was much higher than in theirs (774 vs.\ 121) and the number of items was also higher (60/120 vs.\ 40). 
We will revisit the absence of syntactic interference in our data in the general discussion, where we investigate the question using computational modeling.

@@

\subsubsection{Procedure}
The experiment was conducted in two experimental sessions for practical reasons (each of the sessions lasted approximately two hours). Sessions were separated by one to eight weeks for each participant (the median gap was 2 weeks, the first and third quartiles being 1 and 3 weeks). The procedure of both sessions was identical.

@@

\subsubsection{EEG recording and processing}
The EEG was recorded with 24 Ag/AgCl scalp electrodes, positioned according to the international 10-20 system. During recording, an electrode at the left mastoid was used as reference and AFz as ground. The sampling rate was 500 Hz. Eye-movements were monitored with six electrodes which were positioned above, below and at the outer canthus of both eyes. Impedances of all electrodes were kept below 5k$\Omega$.

@@

Comprehension accuracy was reduced for high vs.\ low semantic interference conditions. There was no indication that syntactic interference affected accuracy, although the sign of the effect was in the expected direction, but  amounting to a 1\% difference on the probability scale between the high and low syntactic interference conditions. There was a numerical suggestion of an interaction: the difference in accuracy between low and high semantic interference conditions was larger when syntactic interference was high vs.\ when it was low (17 vs.\ 8 \% difference).

@@

\textcite{mertzen} presented simulations that show approximately equally sized interference effects 
for both the syntactic and semantic manipulations. Their model assumed that three cues were used at the verb: \{$\pm$ grammatical subject\}, \{$\pm$ same clause\} and \{$\pm$ animacy\}. The upper parts of Figure \ref{fig:model_predictions} a and b reproduce the \textcite{mertzen} simulations. \ref{fig:model_predictions_rts} shows the prior predictive distributions of by-condition retrieval times that were generated by sampling from a prior distribution ($Beta(a=4,b=6)$) for the free scaling parameter $F$ and then generating data from the model while holding all other parameters in the model at fixed values. The $F$ parameter scales the activation to a retrieval time. The prior $Beta(a=4,b=6)$ was chosen in earlier work \parencite{jaeger_etal_2020,vasishth2020_abc}. Figure \ref{fig:model_predictions_effects} shows the prior predictive distributions of the effects of interest.

@@

Could the absence of syntactic interference be reconciled with a cue-based retrieval account? We speculated that one possibility might be that the parser tracks hierarchical structure and that this hierarchical information is used when searching for a noun in memory. Specifically, the relevant syntactic cues that might be used at the verb are not  \{$\pm$ grammatical subject\} and \{$\pm$ same clause\} but rather a composite cue \{$\pm$ subject-in-same-clause\}. This proposal is in the spirit of the  \textcite{dillon2013} and \textcite{Sturt2003} claim, that the parser intelligently targets only the syntactically relevant noun during antecedent retrieval.  In the present case, it is possible that syntactically relevant noun is only the one in the same clause as the verb.

@@

The \citet{Lewis2005} model is generally used to predict reading times. The scaling factor $F$ links the activation of items in memory to retrieval times \citep{Lewis2005,anderson_etal_2004_ACT_R}. But this scaling factor can be adjusted to link activation to brain responses to use the model to generate predictions on the microvolt scale for ERP data. The modeling reported here is, to the best of our knowledge, the first time that the \citet{Lewis2005} model has been fit quantitatively to ERP data. We sampled the free scaling parameter $F$ from a uniform prior distribution ($Uniform(0.005, 0.04)$). This prior reflects the a priori belief that effects will range from 0.5 to 4 $\mu V$ with equal likelihood for all effect sizes in this range. This uninformative prior was chosen because there exists no previously published ERP data on the Van Dyke (2007) design. We are modeling N400 amplitude differences, so larger predicted effect sizes correspond to a more negative N400. To reflect this, the predicted effects were multiplied with $-1$. As in the reading time modeling, to avoid overfitting we fitted $F$ only to the semantic interference effect in order to generate posterior predictive distributions of the syntactic and semantic interference effects. For the ERP modeling, 473 out of 20,000 sampled values of $F$ lead to simulated effects within the range of the observed effect and were thus used for posterior predictions. The prior and posterior predictions for the effects of interest on the microvolt scale in comparison to our ERP data are shown in Figure \ref{fig:model_predictions_ERP}. The prior and posterior predictions of the model overlap with the effects of interest in the data. The general effect pattern (no syntactic interference effect, but a semantic interference effect) is similar in both the prior and posterior predictive distributions. Therefore, the assumption that the relevant syntactic cue is \{$\pm$ subject-in-same-clause\} cue, rather than the \{$\pm$ grammatical subject\} cue originally assumed in previous modeling work, seems reasonable for the present ERP study as well.

@@

There are of course important examples of the parser ignoring syntactic information. Three dramatic examples are the grammaticality illusion in multiple embeddings in English \parencite{gibsonthomas99}, syntactic local coherence effects \parencite{taboretal04}, and the agreement attraction phenomenon across different languages \parencite[e.g.,][]{wagersetal,lago_etal_2021,tucker2015representing}.  However, there exists  evidence inconsistent with these findings. Regarding grammaticality illusions in double-center embeddings, the grammaticality illusion does not occur in German and Dutch \parencite{VSLK11,FrankEtAl2015}. Regarding syntactic local coherence effects, a large-sample study \parencite{lcpaape2023} presents evidence against syntactic local coherence. Regarding agreement attraction, there seems to be at least one language with rich case marking (Czech) that does not show any evidence of agreement attraction \parencite{chromy2023number}. Thus, it seems that, cross-linguistically, syntax does generally play a central role in building incremental structure and in completing dependencies. Of course, this does not rule out the possibility that, under certain conditions, other factors dominate.

>>>>

Reading studies have been the dominant method used to investigate retrieval interference \parencite[see the review in][]{jaeger_etal_2017}.  We aimed to investigate the role of syntactic and semantic cues during sentence comprehension with higher-powered experiments using  self-paced reading (SPR) as well as electroencephalography (EEG), which to our knowledge has never been used to investigate interference during subject-verb dependency resolution.

@@

The SPR experiments were conducted web-based via Prolific and the EEG experiment was of course run in the lab.  Because the signal-to-noise ratio of EEG data requires a large number of trials per condition and participant, we constructed 120 items. In order to keep the SPR and EEG studies comparable, we also used 120 items in the SPR experiments. In both methods, as it would have been very taxing for the participants to read all 120 items in one experimental session, we split the experiments into two sessions.

The SPR study, hereafter Experiment 1a, was initially designed so that participants read items in the first session (1-60) and then, after a gap of 1-20 days, they read the the second set of items (61-120). The variation in the gap is due to the fact that we could not control when the participants would do the second session; however, the mean and median gaps were  about 11 days, with the first and third quartiles being 9 and 12 days.

@@

The Bayes factor is analogous to the frequentist ANOVA or likelihood ratio test: it compares the marginal likelihood of a full model to that of a baseline null model; this is often written as $BF_{10}$, with the subscripts indicating that marginal likelihood of the full model is in the numerator and that of the null model is in the denominator. One major advantage of the Bayes factor over the frequentist analogue is that it takes the uncertainty of the parameters into account. This leads to more conservative inferences compared to frequentist ANOVA, which only takes the maximum likelihood estimate of the parameter into account \parencite[see][for detailed discussion]{schad_etal_2022_BF}. Another important advantage of the Bayes factor -- one that is very relevant for the present work -- is that it is possible to find evidence for or against an effect. This stands in contrast to the frequentist ANOVA which, in its standard usage, is designed to only furnish evidence against the null.

@@

\subsubsection{Comparison with previous findings}
The results of the present study stand in contrast to previous findings \citep{vandyke07, mertzen}. The present study is the only one to find semantic and no syntactic interference effects. The difference in observed effects could be explained by differences in statistical power or by methodological differences. The present study has higher statistical power than the previous studies. \citet{vandyke07} had 35 - 40 participants and 36 - 48 items in each of her three experiments, respectively. \textcite{mertzen} tested 61 English speakers and 121 German speakers with 40 items each. In contrast, we tested 774 participants with at least 60 items each (a subset of 160 participants read all 120 items). Therefore, the present study is the study with the highest power to date using this 2 $\times$ 2 interference design. Consequently, the estimates from the present study are the most precise ones so far for this design. The smaller effect size in the present study is a consequence of higher power, and the magnitude of the effects observed is comparable to that of other large-scale reading studies \citep{nicenboim} and meta-analyses on interference \citep{jaeger_etal_2017}.

An obvious difference from previous studies is the experimental methodology. The majority of the previous experiments employed eye-tracking while reading, while we used a self-paced reading task. Self-paced reading (SPR) might have a greater tendency to be affected by interference compared to natural reading (as in eye-tracking studies), due to the restricted reading format in SPR: Comprehenders have no opportunity to revisit previous material during conventional self-paced reading \citep[but see][]{BSPR}. SPR is likely to be more demanding on the comprehender's memory, which might make it easier to detect interference effects. This argumentation is supported by our estimation of the results of \citeauthor{vandyke07}'s (\citeyear{vandyke07}) Experiment 1, the only other SPR experiment using this design: 
In the critical region, the estimates of both main effects (syntactic and semantic interference) are very similar to the results of the present study; the interaction effect observed by \citeauthor{vandyke07} in the spill-over region is also similar to our estimate. However, it remains unclear why semantic interference was especially prominent in our study compared to the previous ones.

The contrast between our results and those from  \citeauthor{mertzen}'s (\citeyear{mertzen}) German experiment is surprising given that both used German items and partially even the same linguistic material. So, why did we find a semantic interference effect in all regions, whereas \citet{mertzen} only found one in the pre-critical region? Apart from the methodological difference (eye-tracking vs.\ self-paced reading), a plausible  explanation for the divergence is lower power in \citeauthor{mertzen}'s (\citeyear{mertzen}) study. While they already had considerably more participants and items than many psycholinguistic studies, in the present study the number of participants was much higher than in theirs (774 vs.\ 121) and the number of items was also higher (60/120 vs.\ 40). 
We will revisit the absence of syntactic interference in our data in the general discussion, where we investigate the question using computational modeling.

@@

\subsubsection{Procedure}
The experiment was conducted in two experimental sessions for practical reasons (each of the sessions lasted approximately two hours). Sessions were separated by one to eight weeks for each participant (the median gap was two weeks, the first and third quartiles being one and three weeks). The procedure of both sessions was identical.

@@

\subsubsection{EEG recording and processing}
The EEG was recorded with 24 Ag/AgCl scalp electrodes, positioned according to the international 10-20 system. During recording, an electrode at the left mastoid was used as reference and AFz as ground. The sampling rate was 500 Hz. Eye-movements were monitored with six electrodes which were positioned above, below and at the outer canthus of both eyes. Impedances of all electrodes were kept below 5\,k$\Omega$.

@@

Comprehension accuracy was reduced for high vs.\ low semantic interference conditions. There was no indication that syntactic interference affected accuracy. The sign of the effect was in the expected direction, but  amounting to a 1\% difference on the probability scale between the high and low syntactic interference conditions. There was a numerical suggestion of an interaction: the difference in accuracy between low and high semantic interference conditions was larger when syntactic interference was high vs.\ when it was low (17 vs.\ 8 \% difference).

@@

\textcite{mertzen} presented simulations that show approximately equally sized interference effects 
for both the syntactic and semantic manipulations. Their model assumed that three cues were used at the verb: \{$\pm$ grammatical subject\}, \{$\pm$ same clause\} and \{$\pm$ animacy\}. The upper parts of Figure \ref{fig:model_predictions} a and b reproduce the \textcite{mertzen} simulations. Figure \ref{fig:model_predictions_rts} shows the prior predictive distributions of by-condition retrieval times that were generated by sampling from a prior distribution ($Beta(a=4,b=6)$) for the free scaling parameter $F$ and then generating data from the model while holding all other parameters in the model at fixed values. The $F$ parameter scales the activation to a retrieval time. The prior $Beta(a=4,b=6)$ was chosen in earlier work \parencite{jaeger_etal_2020,vasishth2020_abc}. Figure \ref{fig:model_predictions_effects} shows the prior predictive distributions of the effects of interest.

@@

Could the absence of syntactic interference be reconciled with a cue-based retrieval account? We speculate that one possibility might be that the parser tracks hierarchical structure and that this hierarchical information is used when searching for a noun in memory. Specifically, the relevant syntactic cues that might be used at the verb are not  \{$\pm$ grammatical subject\} and \{$\pm$ same clause\} but rather a composite cue \{$\pm$ subject-in-same-clause\}. This proposal is in the spirit of the  \textcite{dillon2013} and \textcite{Sturt2003} claim, that the parser intelligently targets only the syntactically relevant noun during antecedent retrieval.  In the present case, it is possible that syntactically relevant noun is only the one in the same clause as the verb.

@@

The \citet{Lewis2005} model is generally used to predict reading times. The scaling factor $F$ links the activation of items in memory to retrieval times \citep{Lewis2005,anderson_etal_2004_ACT_R}. But this scaling factor can be adjusted to link activation to brain responses to use the model to generate predictions on the microvolt scale for ERP data. The modeling reported here is, to the best of our knowledge, the first time that the \citet{Lewis2005} model has been fit quantitatively to ERP data. We sampled the free scaling parameter $F$ from a uniform prior distribution ($Uniform(0.005, 0.04)$). This prior reflects the a priori belief that effects will range from 0.5 to 4 $\mu V$ with equal likelihood for all effect sizes in this range. This uninformative prior was chosen because there exists no previously published ERP data on the Van Dyke (2007) design. We are modeling N400 amplitude differences, so larger predicted effect sizes correspond to a more negative N400. To reflect this, the predicted effects were multiplied with $-1$. As in the reading time modeling, to avoid overfitting we fitted $F$ only to the semantic interference effect in order to generate posterior predictive distributions of the syntactic and semantic interference effects. For the ERP modeling, 473 out of 20,000 sampled values of $F$ lead to simulated effects within the range of the observed effect and were thus used for posterior predictions. The prior and posterior predictions for the effects of interest on the microvolt scale in comparison to our ERP data are shown in Figure \ref{fig:model_predictions_ERP}. The prior and posterior predictions of the model overlap with the effects of interest in the data. The general effect pattern (no syntactic interference effect, but a semantic interference effect) is similar in both the prior and posterior predictive distributions. Therefore, the assumption that the relevant syntactic cue is the \{$\pm$ subject-in-same-clause\} cue -- rather than the previously assumed \{$\pm$ grammatical subject\} cue --  seems reasonable for the present ERP study as well.

@@

There are of course important examples of the parser ignoring syntactic information. Three dramatic examples are the grammaticality illusion in multiple embeddings in English \parencite{gibsonthomas99}, syntactic local coherence effects \parencite{taboretal04}, and the agreement attraction phenomenon across different languages \parencite[e.g.,][]{wagersetal,lago_etal_2021,tucker2015representing}.  However, there exists  evidence inconsistent with these findings. Regarding grammaticality illusions in double-center embeddings, the grammaticality illusion does not occur in German and Dutch \parencite{VSLK11,FrankEtAl2015}. Regarding syntactic local coherence effects, a large-sample study \parencite{lcpaape2023} presents evidence against syntactic local coherence. Regarding agreement attraction, there seems to be at least one language with rich case marking (Czech) that does not show any evidence of number agreement attraction \parencite{chromy2023number}. Thus, it seems that, cross-linguistically, syntax does generally play a central role in building incremental structure and in completing dependencies. Of course, this does not rule out the possibility that, under certain conditions, other factors dominate.
